Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-20 parallel.tex
added GPU section, including a new figure.  Also include the script that generated the figure.

>>>>

\subsection{GPU parallelization and scaling}

Many of today's leading supercomputers use a heterogeneous structure:
on a single node of a distributed-memory platform, a multi-core CPU is
often paired with one or more many-core accelerators.  One programming
model that has been shown to successfully take advantage of these
hardware accelerators is to run the serial component of the algorithm
on the CPU, and the vector-parallel part of the algorithm on the
many-core hardware accelerator. Due to the much higher computational
performance of the many-core accelerators, if a code can be ported to
effectively utilize this heterogeneous architecture, massive speedup
in simulation performance may be achieved.  This heterogeneous trend
is likely to continue in future supercomputers, due to the relatively
low energy needs of hardware accelerators as well as other
factors. Thus, to ensure \enzo's efficiency on future high-end
computational platforms, some of the most time-consuming parts of
\enzo\ have been ported to many-core architecture.  Since GPUs are
currently the most popular many-core accelerator, NVIDIA's CUDA C/C++
programming technology was chosen to port some \enzo\ modules to
NVIDIA GPUs.

Thus far, the PPM and Dedner MHD solvers (described in
Sections~\ref{sec.hydro.ppm} and~\ref{sec.num.hydro-muscl},
respectively), have been ported to take advantage of GPUs using CUDA.
Since the CPU and GPU on a node have access to separate pools of RAM,
fields updated by other modules will be transferred to the GPU before
calling the GPU fluid solver.  Correspondingly, after calling the GPU
solver, the updated fields will be transferred back to CPU. This
ensures that the GPU-parallelized solvers can work correctly with
other parts of the code that have not yet been ported to the GPU, as
well as with the communication infrastructure within \enzo.  In
addition, the GPU solver supports SAMR, which is not a trivial task.
As discussed in Section~\ref{sec.amr}, one of the key steps in \enzo's
AMR implementation is flux correction, which is required when each
level of resolution is allowed to take its own time step.  In the GPU
version of the solvers, the fluxes are calculated on the GPU and only
the fluxes required for flux correction are transferred back to the
CPU.  This reduces data transfer overhead, which can be substantial in
a heterogeneous architecture of this sort.

The key step in porting to many-core architectures such as the GPU is
exposing the massive parallelism inherent in the algorithm. Due to the
explicit, directionally-split stencil pattern of both the PPM and
Dedner MHD solvers, they are inherently massively parallel and thus
should be good fit for hardware acceleration. Both solvers essentially
contain basically two parts -- computation of fluxes, and cell update.
When computing fluxes, the basic procedure is to compute the flux at
each cell-interface given the inputs from neighboring cells. In the
cell update part, the fundamental computation is updating the
cell-centered values using the previously-computed fluxes. In the CPU
serial code, a loop over the grid is used, where the loop body
contains these basic computations. In both parts,
algorithmically-different loop iterations are completely independent
of each other. Thus, the parallelization scheme is naturally mapping
one GPU thread to one iteration of the loop. However, the original CPU
code, which operates on each grid as a serial process, does not
completely expose this parallelism as some small temporary arrays are
reused among loop iterations.  This re-use of arrays introduces data
dependency among loop iterations, which is undesirable for GPU
parallelization.  Because of this re-use of arrays, the main change in
porting the serial CPU solvers to allow massively parallel GPU
computation was replacing those shared temporary arrays by larger
temporary arrays that are not shared among loop iterations. This
change exposed the massive parallelism in the algorithm, which can
then be accelerated in a straightforward manner using CUDA.

To illustrate the speedup provided by porting two solvers to GPUs, we
show the results of a weak scaling test of driven turbulence in
Figure~\ref{fig:gpu_scaling}.  This \enzo\ problem type uses the
Dedner MHD solver, which runs on both CPUs and GPUs, and has been
tuned to maximize performance on both platforms.  The benchmarking platform is a Cray
XK6 system where each node has a 16-core AMD Opteron 6272 CPU and a single
NVIDIA Tesla K20 GPU, which has 2,496 CUDA cores and a maximum
theoretical speed of 1.17 Tflops when used in double precision.
For the CPU run, all the 16 cores in each node
are used by launching 16 MPI processes per node. For the GPU run, 16 MPI
processes are also launched per node, and all processes on this node share the single GPU
on that node. This can work because NVIDIA's MPS (Multi-Process
Service) technology allows multiple processes to concurrently use the
same GPU.  A series of weak scaling tests were run, varying from 1
to 8 nodes (16 to 128 MPI processes), with the number of cells per
node fixed to $256^3$.  Figure~\ref{fig:gpu_scaling} displays the
scaling results in terms of cell updates per second per node -- as a
result, ideal weak scaling is a horizontal line.  In general, the
simulations using the GPU-accelerated MHD solver perform roughly 5
times better overall on this system than the CPU-only calculations.

\begin{figure}
\centerline{\includegraphics[width=0.5\textwidth]{figures/gpu_scaling.eps}}
\caption{Weak scaling performance of MHD turbulence using the Dedner
MHD algorithm, which has both CPU and GPU solvers.  Each node of the
machine (a Cray XK6) contains one 16-core AMD Opteron 6272 CPU and 1
NVIDIA Tesla K20 GPU.  In this test, there are $256^3$ cells/node.
The y-axis shows cell updates per second per node, which for ideal
weak scaling is a horizontal line.  Blue line: CPU solver.  Red line:
GPU solver.}
\label{fig:gpu_scaling}
\end{figure}
